---
title: "EDA notebook"
author: "Sonia Dalal and Sanjana Malipeddi"
date: "2024-04-21"
output: html_document
---
## Loading Packages
```{r}



library(readr)
library(tibble)
library(dplyr)
library(tidyverse)
library(ggplot2)
data_replic <- read_csv("/Users/soniadalal/Desktop/302W/302W Project/data/data and replication code/data_replic.csv")
head(data_replic)
summary(data_replic)
```


## This dataset shows a comprehensive set of variables with demographic, economic, and crime-related metrics across different countries over time. The UNODC recorded homicide rates and deportations of convicts can be used to determine the impact of deportations on crime rates in different regions of home countries. This relationship can also be analyzed through a socioeconomic lense using the income data provided

iso3: ISO 3166-1 alpha-3 country code representing each country.
year: Year of the data record.
region: World Bank-defined global region that the country belongs to.
income_group: World Bank-defined income category of the country (e.g., Low income, Upper middle income).
homrates_unodc: Homicide rates per 100,000 population as reported by the United Nations Office on Drugs and Crime.
homrates_who: Homicide rates per 100,000 population as reported by the World Health Organization.
deport_convict: Deportation rates of convicted individuals per 100,000 population.
deport_convict_d: Differenced series of deportation rates of convicted individuals, indicating change from the previous year.
deport_convict_lead: Lead series of deportation rates of convicted individuals, indicating the value for the forthcoming year.
deport_nonconvict: Deportation rates of non-convicted individuals per 100,000 population.
corruption: Higher corruption index indicates lower rates of corruption
instr_enforce_everify_lag1: Previous year's enforcement measure related to electronic verification systems.
polity2: Score from the Polity IV project measuring the level of democracy and autocracy.
pop_sh14: Percentage of the population aged under 14 years.
population_log: Logarithm of the population size to normalize the data.
remittances_GDP: Remittances received as a percentage of the GDP.
urban_interp: Urbanization rate interpolated for non-census years.
war_intrastate: Indicator of whether there was an intrastate war occurring.
yr_sch: Average years of schooling in the population.

# Remove rows with missing values in key columns, remove NA columns.
```{r}

df <- data_replic %>%
  select(-...19, -...23)

newdf <- data_replic %>%
  select(-...19, -...23) %>%
  filter(!is.na(deport_nonconvict) & !is.na(homrates_unodc) & !is.na(region) & !is.na(deport_convict) & !is.na(income_group)) 
newdf
```
#Make a correlation plot
```{r}
install.packages("corrplot")
library(corrplot)

```
```{r}
df <- df %>%
  mutate(
    income_group_numeric = case_when(
      income_group == "Low income" ~ 1,
      income_group == "Lower middle income" ~ 2,
      income_group == "Upper middle income" ~ 3,
      income_group == "High income: nonOECD" ~ 4,
      income_group == "High income: OECD" ~ 5,
      TRUE ~ NA_real_  # Handles any cases that don't match the above
    )
  )
```
# General overview of correlations between different variables
```{r}
# Select relevant variables and remove NA values for correlation analysis
cor_data <- df %>%
  select(corruption, homrates_unodc, GDPpercapita_const2010USD_log, urban_interp, year, deport_convict, deport_nonconvict, polity2, yr_sch, income_group_numeric) %>%
  na.omit()  # Remove rows with NA values

# Calculate the correlation matrix
cor_matrix <- cor(cor_data)
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         addCoef.col = "black",  # Add correlation coefficients to enhance readability
         tl.col = "black", tl.srt = 45,  # Adjust text label color and rotation for better visibility
         diag = FALSE,  # Exclude diagonal elements to avoid redundancy
         number.cex = 0.8,  # Adjust size of correlation coefficients
         tl.cex = 0.8)
```
Shows that as deportation increases, homicide rates increase but over the years available in the data, homicide rates and deportations rates did not increase significantly.

#Income distribution Counts in dataset for each region
```{r}
# Bar plot
ggplot(df, aes(x = region, fill = income_group)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Income Distribution Counts for Each Region")
```
More data on income group is available for Latin America & Caribbean and Europe and Central Asia than other regions.
#Convict Deportation distribution in dataset by Region
```{r}
# Bar plot
ggplot(df, aes(x = region, fill = deport_convict)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Convict Deportation Rate Distribution Counts per Region")
```
More data on convict deportation is available for Latin America & Caribbean and Europe and Central Asia than other regions.
#Non-Convict Deportation distribution in dataset by Region
```{r}
# Bar plot
ggplot(df, aes(x = region, fill = deport_nonconvict)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Non-Convict Deportation Rate Distribution Counts per Region")
```

More data on deportation of nonconvicts is available for Latin America & Caribbean and Europe and Central Asia than other regions.
#Corruption distribution in dataset by Region
```{r}
# Bar plot
ggplot(df, aes(x = region, fill = corruption)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Corruption Score Distribution Counts per Region")
```
More data on corruption is available for Latin America & Caribbean and Europe and Central Asia than other regions.
# Boxplot of Corruption Scores for all the Regions
```{r}
# Filter the dataset to include only relevant variables

corruption_data <- data_replic %>%
  select(region, corruption)

# Summary statistics by region
summary_stats <- corruption_data %>%
  group_by(region) %>%
  summarise(mean_corruption = mean(corruption, na.rm = TRUE),
            median_corruption = median(corruption, na.rm = TRUE),
            min_corruption = min(corruption, na.rm = TRUE),
            max_corruption = max(corruption, na.rm = TRUE))

# Visualization: Box plot of corruption scores by region
boxplot_corruption <- ggplot(corruption_data, aes(x = region, y = corruption, fill = region)) +
  geom_boxplot() +
  labs(title = "Corruption Scores by Region",
       x = "Region",
       y = "Corruption Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# Display the summary statistics and box plot
print(summary_stats)
print(boxplot_corruption)


```

Latin America & Caribbean have more corruption on average than Europe & Central Asia
#Income distribution proportions for each region in data with no NA for deportation or homicides.
```{r}
# Calculate the count per income group within each region
newdf_with_count <- newdf %>%
  group_by(region, income_group) %>%
  summarise(count = n(), .groups = 'drop')
new
# Calculate the total count per region
totals <- newdf_with_count %>%
  group_by(region) %>%
  summarise(total = sum(count), .groups = 'drop')

# Join the totals back to the original data frame
newdf_with_proportions <- newdf_with_count %>%
  left_join(totals, by = "region") %>%
  mutate(proportion = count / total)

# Plot the proportions instead of count
ggplot(newdf_with_proportions, aes(x = region, y = proportion, fill = income_group)) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion")
```
Europe & Central Asia has a higher proportion of High income whereas Latin America & Caribbean has a higher proportion of Upper middle income.

#Income distribution proportions for each region with darker colors corresponding to lower income
```{r}
# Order the income_group factor based on income levels from highest to lowest
newdf_with_proportions$income_group <- factor(
  newdf_with_proportions$income_group, 
  levels = c(
    "Low income",
    "Lower middle income",
    "Upper middle income",
    "High income: nonOECD",
    "High income: OECD" 
    
  ),
  ordered = TRUE
)
ggplot(newdf_with_proportions, aes(x = region, y = proportion, fill = income_group)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single')) +
  scale_fill_brewer(palette = "Blues", direction = -1) + # Use reversed Blues palette for darker colors on lower incomes
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Proportion", fill = "Income Group")

```
# Correlational plot with Convict Deportation Rates and Homicide Rates by Region (Scaled and Unscaled)
```{r}
ggplot(newdf, aes(x = deport_convict, y = homrates_unodc)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region, scale = "free") +
  labs(title = "Correlation Between Convict Deportation Rates and Homicide Rates by Region",
       x = "Convict Deportation Rate",
       y = "UNODC Homicide Rate")

ggplot(newdf, aes(x = deport_convict, y = homrates_unodc)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region) +
  labs(title = "Correlation Between Convict Deportation Rates and Homicide Rates by Region",
       x = "Non-Convict Deportation Rate",
       y = "UNODC Homicide Rate")
correlation_by_region <- newdf %>%
  group_by(region) %>%
  summarize(correlation = cor(deport_convict, homrates_unodc, use = "complete.obs"))

print(correlation_by_region)
```
# Correlational plot with Non-Convict Deportation Rates and Homicide Rates by Region (Scaled and Unscaled)
```{r}
ggplot(newdf, aes(x = deport_nonconvict, y = homrates_unodc)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region, scale = "free") +
  labs(title = "Correlation Between Non-Convict Deportation Rates and Homicide Rates by Region",
       x = "Non-Convict Deportation Rate",
       y = "UNODC Homicide Rate")

ggplot(newdf, aes(x = deport_nonconvict, y = homrates_unodc)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region) +
  labs(title = "Correlation Between Non-Convict Deportation Rates and Homicide Rates by Region",
       x = "Non-Convict Deportation Rate",
       y = "UNODC Homicide Rate")

correlation_by_region2 <- newdf %>%
  group_by(region) %>%
  summarize(correlation = cor(deport_nonconvict, homrates_unodc, use = "complete.obs"))

print(correlation_by_region2)
```
These correlational plots show the most comprehensive data for non-convicts and convicts for Latin American and the Carribean. For this reason we will forcus on this region.
# Distribution of Years in the Dataset
```{r}


# Preparing the data by ensuring 'year' is available and correctly formatted
newdf <- mutate(newdf, year = as.integer(year))

# Combine into one dataframe for plotting

# Plotting the histogram of year distributions across datasets
ggplot(newdf, aes(x = year)) +
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) +
  labs(title = "Distribution of Years", x = "Year", y = "Count") +
  scale_fill_manual(values = c("blue", "red", "green")) +
  theme_minimal()

```
Data only available from 2004 to 2014. This plus the correlation plot shows a limitation that there hasn't been much change in the variables which may make it more difficult to find relationships in this time. 

```{r}
# Filter the dataset for Latin America & Caribbean region
latin_america_data <- newdf %>%
  filter(region == "Latin America & Caribbean")

# Impute missing data for odd years using linear interpolation
imputed_data1 <- latin_america_data %>%
  tidyr::complete(year = seq(min(year), max(year), by = 2)) %>%
  mutate(homrates_unodc = approx(year[!is.na(homrates_unodc)], homrates_unodc[!is.na(homrates_unodc)], year)$y)

# Create the line plot
ggplot(imputed_data1, aes(x = year, y = homrates_unodc)) +
  geom_line() +
  labs(title = "Homicide Rates in Latin America & Caribbean",
       x = "Year",
       y = "Homicide Rates (per 100,000 population)") +
  theme_minimal()
```
```{r}


# Filter the dataset for Latin America & Caribbean region
europe_central_asia_data <- newdf %>%
  filter(region == "Europe & Central Asia")

# Impute missing data for odd years using linear interpolation
imputed_data2 <- europe_central_asia_data %>%
  tidyr::complete(year = seq(min(year), max(year), by = 2)) %>%
  mutate(homrates_unodc = approx(year[!is.na(homrates_unodc)], homrates_unodc[!is.na(homrates_unodc)], year)$y)

# Create the line plot
ggplot(imputed_data2, aes(x = year, y = homrates_unodc)) +
  geom_line() +
  labs(title = "Homicide Rates in Europe and Central Asia",
       x = "Year",
       y = "Homicide Rates (per 100,000 population)") +
  theme_minimal()

```
```{r}
combined_data <- rbind(imputed_data1, imputed_data2)
ggplot(combined_data, aes(x = year, y = homrates_unodc, color = region)) +
  geom_line() +
  labs(title = "Homicide Rates in Latin America & Caribbean vs. Europe & Central Asia",
       x = "Year",
       y = "Homicide Rates (per 100,000 population)") +
  theme_minimal() +
  scale_color_manual(values = c("blue", "red"))
```
Europe and Central Asia Homicide rates are low and slightly decrease over the years while homicide rates in Latin America and the Caribbean are much higher and increase then decrease.
# Make new dataset with just Latin America and the Caribbean
```{r}
# Filter the dataset for Latin America & Caribbean region
latin_america_data <- newdf %>%
  filter(region == "Latin America & Caribbean")

# Impute missing data for odd years using linear interpolation
imputed_data <- latin_america_data %>%
  tidyr::complete(year = seq(min(year), max(year), by = 2)) %>%
  mutate(homrates_unodc = approx(year[!is.na(homrates_unodc)], homrates_unodc[!is.na(homrates_unodc)], year)$y,
         deport_convict = approx(year[!is.na(deport_convict)], deport_convict[!is.na(deport_convict)], year)$y,
         deport_nonconvict = approx(year[!is.na(deport_nonconvict)], deport_nonconvict[!is.na(deport_nonconvict)], year)$y)

# Create the line plot
ggplot(imputed_data, aes(x = year, y = homrates_unodc)) +
  geom_line(aes(color = "Homicide Rates"), size = 1) +
  geom_line(aes(y = deport_convict, color = "Convict Deportation Rates"), linetype = "dashed", size = 1) +
  geom_line(aes(y = deport_nonconvict, color = "Non-Convict Deportation Rates"), linetype = "dotdash", size = 1) +
  labs(title = "Homicide and Deportation Rates in Latin America & Caribbean",
       x = "Year",
       y = "Rates per 100,000 population",
       color = "Legend") +
  scale_color_manual(values = c("Homicide Rates" = "red",
                                "Convict Deportation Rates" = "blue",
                                "Non-Convict Deportation Rates" = "green")) +
  theme_minimal() +
  theme(legend.title = element_blank())

```
```{r}
# Filter data for Europe & Central Asia
europe_data <- newdf %>%
  filter(region == "Europe & Central Asia")

# Impute missing data for odd years using linear interpolation
imputed_europe_data <- europe_data %>%
  tidyr::complete(year = seq(min(year), max(year), by = 2)) %>%
  mutate(homrates_unodc = approx(year[!is.na(homrates_unodc)], homrates_unodc[!is.na(homrates_unodc)], year)$y,
         deport_convict = approx(year[!is.na(deport_convict)], deport_convict[!is.na(deport_convict)], year)$y,
         deport_nonconvict = approx(year[!is.na(deport_nonconvict)], deport_nonconvict[!is.na(deport_nonconvict)], year)$y)

# Create the plot for Europe & Central Asia
plot_europe <- ggplot(imputed_europe_data) +
  geom_line(aes(x = year, y = homrates_unodc, color = "Homicide Rates"), size = 1) +
  geom_line(aes(x = year, y = deport_convict, color = "Convict Deportation Rates"), size = 1, linetype = "dashed") +
  geom_line(aes(x = year, y = deport_nonconvict, color = "Non-Convict Deportation Rates"), size = 1, linetype = "dotdash") +
  labs(title = "Homicide and Deportation Rates in Europe & Central Asia",
       x = "Year",
       y = "Rates per 100,000 population",
       color = "Legend") +
  scale_color_manual(values = c("Homicide Rates" = "red",
                                "Convict Deportation Rates" = "blue",
                                "Non-Convict Deportation Rates" = "green")) +
  theme_minimal() +
  theme(legend.title = element_blank())

# Display the plot for Europe & Central Asia
plot_europe


```
```{r}
latin_america_tot <- df %>%
  filter(region == "Latin America & Caribbean")

imputed_latin_america_tot <- latin_america_tot %>%
  tidyr::complete(year = seq(min(year), max(year), by = 2)) %>%
  mutate(deport_convict = approx(year[!is.na(deport_convict)], deport_convict[!is.na(deport_convict)], year)$y,
         deport_nonconvict = approx(year[!is.na(deport_nonconvict)], deport_nonconvict[!is.na(deport_nonconvict)], year)$y)

ggplot(imputed_latin_america_tot) +
  geom_line(aes(x = year, y = deport_convict, color = "Convict Deportation Rates"), size = 1) +
  geom_line(aes(x = year, y = deport_nonconvict, color = "Non-Convict Deportation Rates"), size = 1) +
  labs(title = "Deportation Rates in Latin America & Caribbean",
       x = "Year",
       y = "Rates per 100,000 population",
       color = "Legend") +
  scale_color_manual(values = c("Convict Deportation Rates" = "blue",
                                "Non-Convict Deportation Rates" = "green")) +
  theme_minimal() +
  theme(legend.title = element_blank())
```
```{r}
europe_tot <- df %>%
  filter(region == "Europe & Central Asia")

imputed_europe_tot <- europe_tot %>%
  tidyr::complete(year = seq(min(year), max(year), by = 2)) %>%
  mutate(deport_convict = approx(year[!is.na(deport_convict)], deport_convict[!is.na(deport_convict)], year)$y,
         deport_nonconvict = approx(year[!is.na(deport_nonconvict)], deport_nonconvict[!is.na(deport_nonconvict)], year)$y)

ggplot(imputed_europe_tot) +
  geom_line(aes(x = year, y = deport_convict, color = "Convict Deportation Rates"), size = 1) +
  geom_line(aes(x = year, y = deport_nonconvict, color = "Non-Convict Deportation Rates"), size = 1) +
  labs(title = "Deportation Rates in Latin America & Caribbean",
       x = "Year",
       y = "Rates per 100,000 population",
       color = "Legend") +
  scale_color_manual(values = c("Convict Deportation Rates" = "blue",
                                "Non-Convict Deportation Rates" = "green")) +
  theme_minimal() +
  theme(legend.title = element_blank())
```
```{r}

# Filter the dataset to include only relevant variables and regions
region_data <- data_replic %>%
  filter(region %in% c("Latin America & Caribbean", "Europe & Central Asia")) %>%
  select(region, corruption, homrates_unodc)

# Visualization: Scatter plot of corruption scores vs. homicide rates by region
scatter_plot <- ggplot(region_data, aes(x = corruption, y = homrates_unodc, color = region)) +
  geom_point() +
  labs(title = "Corruption Scores vs. Homicide Rates",
       x = "Corruption Score",
       y = "Homicide Rate",
       color = "Region") +
  theme_minimal()

# Display the scatter plot
print(scatter_plot)

```
```{r}

ggplot(newdf, aes(x = corruption, y = deport_convict)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region, scale = "free") +
  labs(title = "Correlation Between Corruption Scores and Convict Deportation Rates by Region",
       x = "Corruption Score",
       y = "Convict Deportation Rate")

ggplot(newdf, aes(x = corruption, y = deport_convict)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region) +
  labs(title = "Correlation Between Corruption Scores and Convict Deportation Rates by Region",
       x = "Corruption Score",
       y = "Convict Deportation Rate")
```
```{r}

ggplot(newdf, aes(x = corruption, y = deport_nonconvict)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region, scale = "free") +
  labs(title = "Corruption Scores & Non-Convict Deportation Rates by Region",
       x = "Corruption Score",
       y = "Non-Convict Deportation Rate")

ggplot(newdf, aes(x = corruption, y = deport_nonconvict)) +
  geom_point(aes(color = region), alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ region) +
  labs(title = "Corruption Scores & Non-Convict Deportation Rates by Region",
       x = "Corruption Score",
       y = "Non-Convict Deportation Rate")
```

```{r}
mean_rates <- newdf %>%
  group_by(region) %>%
  summarize(mean_homicide_rate = mean(homrates_unodc, na.rm = TRUE),
            mean_deport_convict = mean(deport_convict, na.rm = TRUE),
            mean_deport_nonconvict = mean(deport_nonconvict, na.rm = TRUE))

ggplot(mean_rates, aes(x = region)) +
  geom_bar(aes(y = mean_homicide_rate), stat = "identity", position = "dodge", fill = "blue") +
  labs(title = "Mean Homicide Rates by Region (2004-2014)",
       x = "Region",
       y = "Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
ggplot(mean_rates, aes(x = region)) +
  geom_bar(aes(y = mean_deport_convict), stat = "identity", position = "dodge", fill = "red") +
  labs(title = "Mean Convict Deportation Rates by Region (2004-2014)",
       x = "Region",
       y = "Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(mean_rates, aes(x = region)) +
  geom_bar(aes(y = mean_deport_nonconvict), stat = "identity", position = "dodge", fill = "green") +
  labs(title = "Mean Non-Convict Rates by Region (2004-2014)",
       x = "Region",
       y = "Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


